---
title: "BDA Project"
author: "Anonymous"
output:
  pdf_document: null
  toc: yes
  html_document:
    df_print: paged
toc_depth: 1
urlcolor: blue
---

```{r setup, include=FALSE}
# This chunk sets echo = TRUE as default, that is print all code.
# knitr::opts_chunk$set can be used to set other notebook generation options, too.
# include=FALSE inside curly brackets makes this block not be included in the pdf.
knitr::opts_chunk$set(echo = TRUE)
```

# Loaded packages

```{r message=FALSE, warning=FALSE}
#install.packages("readxl")
#nstall.packages("DescTools")
library(readxl)
library(tidyr)
library(DescTools)
library(plyr)
library(ggplot2)
library(reshape2)
library(rstan) 
#library(cmdstanr)
rstan_options(auto_write = TRUE)
options(mc.cores = 1)
# library(loo)
library(bayesplot)
library(posterior)
library(knitr)
```

# Clean data

```{r, results='hide'}
setwd("/Users/vilmatiainen/Documents/Aalto bsc/Year 3/BDA/BDA_project")

multiplesheets <- function(fname) {
  
  # getting info about all excel sheets
  sheets <- readxl::excel_sheets(fname)
  tibble <- lapply(sheets, function(x) readxl::read_excel(fname, sheet = x))
  data_frame <- lapply(tibble, as.data.frame)
  
  # assigning names to data frames
  names(data_frame) <- sheets
  
  init_df <<- data_frame
}

# specifying the path name
path <- "/Users/vilmatiainen/Documents/Aalto bsc/Year 3/BDA/BDA_project/Materials_Municipal_Waste_Stream_1960_2018.xlsx"
multiplesheets(path)

# Dataframe for storing the textile data
all_data <- data.frame(matrix(ncol = 0, nrow = 15))

# Add the data to the dataframe
for (name in names(init_df)) {
  df <- init_df[name][[1]]
  no_na <- drop_na(df)
  row_df <- data.frame(no_na[,-1], row.names=no_na[,1])
  new_df <- row_df[grep("Textiles", rownames(row_df)), ]
  t_df <- t(new_df)
  if (ncol(t_df) > 0) {
    colnames(t_df) <- c(name)
    all_data <<- cbind(all_data, t_df)
  }
}

# Clean up row names
rownames(all_data) <-gsub("X","",as.character(rownames(all_data)))

all_data['year'] <- as.numeric(as.character(rownames(all_data)))
all_data['Materials recycled-%'] <- (all_data['Materials recycled']/all_data['Materials generated'])*100
all_data['Material combusted-%'] <- (all_data['Material combusted']/all_data['Materials generated'])*100
all_data['Materials landfilled-%'] <- (all_data['Materials landfilled']/all_data['Materials generated'])*100

rownames(all_data) <- NULL

#Adjust 1950 to be year 0:
all_data['year'] <- all_data['year']-1950
```

# Plot data

```{r data_plot, fig.height=4, fig.width=5, fig.cap="Data."}
plot(x=all_data[['year']], y=all_data[['Material combusted']], type='l')

```

# Define prior parameters

```{r}
# Separate model:
#Estimate intercept means:
#recycled_intercept <- 680000*0.001 #US population in year 0  times 1kg (0.001 tonnes) for clothing waste in a yera per person
#combusted_intercept <- 0
#landfilled_intercept <- 0

#Estimate slope means:
summary_2 <- summary(lm(`Materials recycled` ~ year, data=all_data[(1:6),]))
recycled_slope <- summary_2$coefficients[2,1]

summary_3 <- summary(lm(`Material combusted` ~ year, data=all_data[(1:6),]))
combusted_slope <- summary_3$coefficients[2,1]

summary_4 <- summary(lm(`Materials landfilled` ~ year, data=all_data[(1:6),]))
landfilled_slope <- summary_4$coefficients[2,1]

beta_mean = c(recycled_slope, combusted_slope, landfilled_slope)

sd_mean = c(sd(all_data[1:6,'Materials recycled']), sd(all_data[1:6,'Material combusted']), sd(all_data[1:6,'Materials landfilled']))
# Calculate the factor by how much the standard deviation of generated textiles has shrunk after between 1960-2005 and 2010-2018
factor <- sd(all_data[1:6, 'Materials generated'])/sd(all_data[7:15, 'Materials generated'])
# Divide sd's by the factor to make them more realistic
sd_mean <- sd_mean/factor

# Intercept means for 1950 0:
recycled_intercept <- max(summary_2$coefficients[1,1], 0)
combusted_intercept <- max(summary_3$coefficients[1,1], 0)
landfilled_intercept <- max(summary_4$coefficients[1,1], 0)
alpha_mean = c(recycled_intercept, combusted_intercept, landfilled_intercept)

# Hierarchical model:
hierarchical_intercept_mean <- mean(alpha_mean)
hierarchical_slope_mean <- mean(beta_mean)
hierarchical_sd_mean <- mean(sd_mean)

```

# Fit models

```{r}
# Prepare data
data_hierarchical = list(
  N = nrow(all_data[7:15,]),
  J = ncol(all_data[,2:4]),
  x = all_data[7:15,'year'],
  y = all_data[7:15, 2:4],
  xpred = (2019-1950),
  sd_mean = hierarchical_sd_mean,
  slope_mean = hierarchical_slope_mean,
  intercept_mean = hierarchical_intercept_mean
)

data_separate = list(
  N = nrow(all_data[7:15,]),
  J = ncol(all_data[,2:4]),
  x = all_data[7:15,'year'],
  y = all_data[7:15, 2:4],
  xpred = (2019-1950),
  alpha_mean = alpha_mean,
  beta_mean = beta_mean,
  sds = sd_mean
)


fit_hierarchical <- stan(file = "Hierarchical_project.stan", data = data_hierarchical) #Had to remove the sigma prior and modify the sd priors bc it wasn't converging
fit_separate <- stan(file = "Separate_project.stan", data = data_separate)
monitor(fit_hierarchical)
monitor(fit_separate)
```

# Convergence analysis

## R-hat and ESS

Hierarchical:

```{r results='asis'}
hier_matr = matrix(data=NA, nrow=6, ncol=3)
colnames(hier_matr) = c('Rhat', 'bulk-ESS', 'tail-ESS')
rownames(hier_matr) = c('alpha[1]', 'beta[1]', 'alpha[2]', 'beta[2]', 'alpha[3]', 'beta[3]')
for (i in 1:3) {
  alpha <- paste('alpha[', as.character(i), ']', sep="")
  beta <- paste('beta[', as.character(i), ']', sep="")
  alpha_mat <- extract_variable_matrix(fit_hierarchical, alpha)
  beta_mat <- extract_variable_matrix(fit_hierarchical, beta)
  rhat_alpha <- posterior::rhat(alpha_mat)
  rhat_beta <- posterior::rhat(beta_mat)
  b_ess_alpha <- posterior::ess_bulk(alpha_mat)
  b_ess_beta <- posterior::ess_bulk(beta_mat)
  t_ess_alpha <- posterior::ess_tail(alpha_mat)
  t_ess_beta <- posterior::ess_tail(beta_mat)
  hier_matr[i+(i-1), 1] <- rhat_alpha
  hier_matr[i+(i-1)+1, 1] <- rhat_beta
  hier_matr[i+(i-1), 2] <- b_ess_alpha
  hier_matr[i+(i-1)+1, 2] <- b_ess_beta
  hier_matr[i+(i-1), 3] <- t_ess_alpha
  hier_matr[i+(i-1)+1, 3] <- t_ess_beta
}
kable(hier_matr, caption='Hierarchical model')
```

Hierarchical had good Rhat and ESS from the start.

Separate:

```{r results='asis'}
sep_matr = matrix(data=NA, nrow=6, ncol=3)
colnames(sep_matr) = c('Rhat', 'bulk-ESS', 'tail-ESS')
rownames(sep_matr) = c('alpha[1]', 'beta[1]', 'alpha[2]', 'beta[2]', 'alpha[3]', 'beta[3]')
for (i in 1:3) {
  alpha <- paste('alpha[', as.character(i), ']', sep="")
  beta <- paste('beta[', as.character(i), ']', sep="")
  alpha_mat <- extract_variable_matrix(fit_separate, alpha)
  beta_mat <- extract_variable_matrix(fit_separate, beta)
  rhat_alpha <- posterior::rhat(alpha_mat)
  rhat_beta <- posterior::rhat(beta_mat)
  b_ess_alpha <- posterior::ess_bulk(alpha_mat)
  b_ess_beta <- posterior::ess_bulk(beta_mat)
  t_ess_alpha <- posterior::ess_tail(alpha_mat)
  t_ess_beta <- posterior::ess_tail(beta_mat)
  sep_matr[i+(i-1), 1] <- rhat_alpha
  sep_matr[i+(i-1)+1, 1] <- rhat_beta
  sep_matr[i+(i-1), 2] <- b_ess_alpha
  sep_matr[i+(i-1)+1, 2] <- b_ess_beta
  sep_matr[i+(i-1), 3] <- t_ess_alpha
  sep_matr[i+(i-1)+1, 3] <- t_ess_beta
}
kable(sep_matr, caption='Separate model')
```

For first separate model all Rhat values were \>1.05 and most ESS values were \<100. Because of this, the model was simplified by removing the sigma prior entirely. This fixed the issue.

## HMC diagnostics

```{r}
check_hmc_diagnostics(fit_hierarchical)
```

Hierarchical had divergences. Changing adapt_delta didn't seem to do much, so instead priors were modified. At this point, a mistake in the alpha_sigma and beta_sigma priors was noticed (too high means and std), so that was fixed. Afterwards, no more divergences.

```{r}
check_hmc_diagnostics(fit_separate)
```

# Posterior predictive checks

```{r}
# Materials combusted 2019 prediction
s_yrep_c <- as.matrix(fit_separate, pars = "yrep_c")
s_yrep_r <- as.matrix(fit_separate, pars = "yrep_r")
s_yrep_l <- as.matrix(fit_separate, pars = "yrep_l")
h_yrep_c <- as.matrix(fit_hierarchical, pars = "yrep_c")
h_yrep_r <- as.matrix(fit_hierarchical, pars = "yrep_r")
h_yrep_l <- as.matrix(fit_hierarchical, pars = "yrep_l")
y_c <- all_data[['Material combusted']]
y_r <- all_data[['Materials recycled']]
y_l <- all_data[['Materials landfilled']]
ppc_dens_overlay(y_c[7:15], yrep = s_yrep_c[1:50,])
```

```{r}
ppc_dens_overlay(y_r[7:15], yrep = s_yrep_r[1:50,])
```

```{r}
ppc_dens_overlay(y_l[7:15], yrep = s_yrep_l[1:50,])
```

```{r}
ppc_dens_overlay(y_c[7:15], yrep = h_yrep_c[1:50,])
```

```{r}
ppc_dens_overlay(y_r[7:15], yrep = h_yrep_r[1:50,])
```

```{r}
ppc_dens_overlay(y_l[7:15], yrep = h_yrep_l[1:50,])
```

# Predictive performance assessment

To assess performance, we can fit the model excluding the year 2018, instead adding that year as the predicted year. We can then compare the true values for 2018 to the predicted values.

Fitting the models:

```{r}
# Prepare data
performance_data_hierarchical = list(
  N = nrow(all_data[7:14,]),
  J = ncol(all_data[,2:4]),
  x = all_data[7:14,'year'],
  y = all_data[7:14, 2:4],
  xpred = (2018-1950),
  sd_mean = hierarchical_sd_mean,
  slope_mean = hierarchical_slope_mean,
  intercept_mean = hierarchical_intercept_mean
)

performance_data_separate = list(
  N = nrow(all_data[7:14,]),
  J = ncol(all_data[,2:4]),
  x = all_data[7:14,'year'],
  y = all_data[7:14, 2:4],
  xpred = (2018-1950),
  alpha_mean = alpha_mean,
  beta_mean = beta_mean,
  sds = sd_mean
)


fit_hierarchical_perf <- stan(file = "Hierarchical_project.stan", data = performance_data_hierarchical) #Had to remove the sigma prior and modify the sd priors bc it wasn't converging
fit_separate_perf <- stan(file = "Separate_project.stan", data = performance_data_separate)
```

Results:

```{r}
ypred_r_h <- extract(fit_hierarchical_perf, 'ypred_r')
ypred_c_h <- extract(fit_hierarchical_perf, 'ypred_c')
ypred_l_h <- extract(fit_hierarchical_perf, 'ypred_l')
pred_h <- c(ypred_r_h, ypred_c_h, ypred_l_h)
ypred_r_s <- extract(fit_separate_perf, 'ypred_r')
ypred_c_s <- extract(fit_separate_perf, 'ypred_c')
ypred_l_s <- extract(fit_separate_perf, 'ypred_l')
pred_s <- c(ypred_r_s, ypred_c_s, ypred_l_s)

real_vals <- all_data[all_data$year==(2018 - 1950),]
reals <- c(real_vals[['Materials recycled']], real_vals[['Material combusted']], real_vals[['Materials landfilled']])

hier_pred = matrix(data=NA, nrow=3, ncol=3)
colnames(hier_pred) = c('Materials recycled', 'Materials combusted', 'Materials landfilled')
rownames(hier_pred) = c('Real', 'Predicted mean', 'Predicted sd')
for (i in 1:3) {
  hier_pred[1,i] <- reals[[i]]
  hier_pred[2,i] <- mean(pred_h[[i]])
  hier_pred[3,i] <- sd(pred_h[[i]])
}
kable(hier_pred, caption='Hierarchical model')
```

```{r}
sep_pred = matrix(data=NA, nrow=3, ncol=3)
colnames(sep_pred) = c('Materials recycled', 'Materials combusted', 'Materials landfilled')
rownames(sep_pred) = c('Real', 'Predicted mean', 'Predicted sd')
for (i in 1:3) {
  sep_pred[1,i] <- reals[[i]]
  sep_pred[2,i] <- mean(pred_s[[i]])
  sep_pred[3,i] <- sd(pred_s[[i]])
}
kable(sep_pred, caption='Separate model')
```

# Sensitivity analysis

```{r}
# Fit models with uniform prior
uniform_hierarchical <- stan(file = "Uniform_hierarchical.stan", data = data_hierarchical)
uniform_separate <- stan(file = "Uniform_separate.stan", data = data_separate)

# Fit models with other prior combo
second_hierarchical <- stan(file = "2ndprior_hierarchical.stan", data = data_hierarchical)
second_separate <- stan(file = "2ndprior_separate.stan", data = data_separate)

hier_sens_df <- as.data.frame(colMeans(as.data.frame(extract(uniform_hierarchical, 'yrep_r'))))
rownames(hier_sens_df) <- NULL
colnames(hier_sens_df) <- c('value')
hier_sens_df['Prior'] <- 'Uniform'
hier_sens_df['x'] <- as.numeric(rownames(hier_sens_df))
temp_df <- as.data.frame(colMeans(as.data.frame(extract(second_hierarchical, 'yrep_r'))))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Second'
temp_df['x'] <- as.numeric(rownames(temp_df))
hier_sens_df <- rbind(hier_sens_df, temp_df)
temp_df <- as.data.frame(colMeans(as.data.frame(extract(fit_hierarchical, 'yrep_r'))))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Original'
temp_df['x'] <- as.numeric(rownames(temp_df))
hier_sens_df <- rbind(hier_sens_df, temp_df)
hier_sens_df['y_type'] <- 'Replicate'

temp_2 <- as.data.frame(extract(uniform_hierarchical, 'ypred_r'))
rownames(temp_2) <- NULL
colnames(temp_2) <- c('value')
temp_2['Prior'] <- 'Uniform'
temp_2['x'] <- as.numeric(rownames(temp_2))
temp_df <- as.data.frame(extract(second_hierarchical, 'ypred_r'))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Second'
temp_df['x'] <- as.numeric(rownames(temp_df))
temp_2 <- rbind(temp_2, temp_df)
temp_df <- as.data.frame(extract(fit_hierarchical, 'ypred_r'))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Original'
temp_df['x'] <- as.numeric(rownames(temp_df))
temp_2 <- rbind(temp_2, temp_df)
temp_2['y_type'] <- 'Prediction'
hier_sens_df <- rbind(hier_sens_df, temp_2)
hier_sens_df['col_val'] <- 'Recycled'

temp_3 <- as.data.frame(colMeans(as.data.frame(extract(uniform_hierarchical, 'yrep_c'))))
rownames(temp_3) <- NULL
colnames(temp_3) <- c('value')
temp_3['Prior'] <- 'Uniform'
temp_3['x'] <- as.numeric(rownames(temp_3))
temp_df <- as.data.frame(colMeans(as.data.frame(extract(second_hierarchical, 'yrep_c'))))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Second'
temp_df['x'] <- as.numeric(rownames(temp_df))
temp_3 <- rbind(temp_3, temp_df)
temp_df <- as.data.frame(colMeans(as.data.frame(extract(fit_hierarchical, 'yrep_c'))))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Original'
temp_df['x'] <- as.numeric(rownames(temp_df))
temp_3 <- rbind(temp_3, temp_df)
temp_3['y_type'] <- 'Replicate'

temp_2 <- as.data.frame(extract(uniform_hierarchical, 'ypred_c'))
rownames(temp_2) <- NULL
colnames(temp_2) <- c('value')
temp_2['Prior'] <- 'Uniform'
temp_2['x'] <- as.numeric(rownames(temp_2))
temp_df <- as.data.frame(extract(second_hierarchical, 'ypred_c'))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Second'
temp_df['x'] <- as.numeric(rownames(temp_df))
temp_2 <- rbind(temp_2, temp_df)
temp_df <- as.data.frame(extract(fit_hierarchical, 'ypred_c'))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Original'
temp_df['x'] <- as.numeric(rownames(temp_df))
temp_2 <- rbind(temp_2, temp_df)
temp_2['y_type'] <- 'Prediction'
temp_3 <- rbind(temp_3, temp_2)
temp_3['col_val'] <- 'Combusted'
hier_sens_df <- rbind(hier_sens_df, temp_3)

temp_3 <- as.data.frame(colMeans(as.data.frame(extract(uniform_hierarchical, 'yrep_l'))))
rownames(temp_3) <- NULL
colnames(temp_3) <- c('value')
temp_3['Prior'] <- 'Uniform'
temp_3['x'] <- as.numeric(rownames(temp_3))
temp_df <- as.data.frame(colMeans(as.data.frame(extract(second_hierarchical, 'yrep_l'))))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Second'
temp_df['x'] <- as.numeric(rownames(temp_df))
temp_3 <- rbind(temp_3, temp_df)
temp_df <- as.data.frame(colMeans(as.data.frame(extract(fit_hierarchical, 'yrep_l'))))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Original'
temp_df['x'] <- as.numeric(rownames(temp_df))
temp_3 <- rbind(temp_3, temp_df)
temp_3['y_type'] <- 'Replicate'

temp_2 <- as.data.frame(extract(uniform_hierarchical, 'ypred_l'))
rownames(temp_2) <- NULL
colnames(temp_2) <- c('value')
temp_2['Prior'] <- 'Uniform'
temp_2['x'] <- as.numeric(rownames(temp_2))
temp_df <- as.data.frame(extract(second_hierarchical, 'ypred_l'))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Second'
temp_df['x'] <- as.numeric(rownames(temp_df))
temp_2 <- rbind(temp_2, temp_df)
temp_df <- as.data.frame(extract(fit_hierarchical, 'ypred_l'))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Original'
temp_df['x'] <- as.numeric(rownames(temp_df))
temp_2 <- rbind(temp_2, temp_df)
temp_2['y_type'] <- 'Prediction'
temp_3 <- rbind(temp_3, temp_2)
temp_3['col_val'] <- 'Landfilled'
hier_sens_df <- rbind(hier_sens_df, temp_3)
```

```{r}
sep_sens_df <- as.data.frame(colMeans(as.data.frame(extract(uniform_separate, 'yrep_r'))))
rownames(sep_sens_df) <- NULL
colnames(sep_sens_df) <- c('value')
sep_sens_df['Prior'] <- 'Uniform'
sep_sens_df['x'] <- as.numeric(rownames(sep_sens_df))
temp_df <- as.data.frame(colMeans(as.data.frame(extract(second_separate, 'yrep_r'))))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Second'
temp_df['x'] <- as.numeric(rownames(temp_df))
sep_sens_df <- rbind(sep_sens_df, temp_df)
temp_df <- as.data.frame(colMeans(as.data.frame(extract(fit_separate, 'yrep_r'))))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Original'
temp_df['x'] <- as.numeric(rownames(temp_df))
sep_sens_df <- rbind(sep_sens_df, temp_df)
sep_sens_df['y_type'] <- 'Replicate'

temp_2 <- as.data.frame(extract(uniform_separate, 'ypred_r'))
rownames(temp_2) <- NULL
colnames(temp_2) <- c('value')
temp_2['Prior'] <- 'Uniform'
temp_2['x'] <- as.numeric(rownames(temp_2))
temp_df <- as.data.frame(extract(second_separate, 'ypred_r'))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Second'
temp_df['x'] <- as.numeric(rownames(temp_df))
temp_2 <- rbind(temp_2, temp_df)
temp_df <- as.data.frame(extract(fit_separate, 'ypred_r'))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Original'
temp_df['x'] <- as.numeric(rownames(temp_df))
temp_2 <- rbind(temp_2, temp_df)
temp_2['y_type'] <- 'Prediction'
sep_sens_df <- rbind(sep_sens_df, temp_2)
sep_sens_df['col_val'] <- 'Recycled'

temp_3 <- as.data.frame(colMeans(as.data.frame(extract(uniform_separate, 'yrep_c'))))
rownames(temp_3) <- NULL
colnames(temp_3) <- c('value')
temp_3['Prior'] <- 'Uniform'
temp_3['x'] <- as.numeric(rownames(temp_3))
temp_df <- as.data.frame(colMeans(as.data.frame(extract(second_separate, 'yrep_c'))))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Second'
temp_df['x'] <- as.numeric(rownames(temp_df))
temp_3 <- rbind(temp_3, temp_df)
temp_df <- as.data.frame(colMeans(as.data.frame(extract(fit_separate, 'yrep_c'))))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Original'
temp_df['x'] <- as.numeric(rownames(temp_df))
temp_3 <- rbind(temp_3, temp_df)
temp_3['y_type'] <- 'Replicate'

temp_2 <- as.data.frame(extract(uniform_separate, 'ypred_c'))
rownames(temp_2) <- NULL
colnames(temp_2) <- c('value')
temp_2['Prior'] <- 'Uniform'
temp_2['x'] <- as.numeric(rownames(temp_2))
temp_df <- as.data.frame(extract(second_separate, 'ypred_c'))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Second'
temp_df['x'] <- as.numeric(rownames(temp_df))
temp_2 <- rbind(temp_2, temp_df)
temp_df <- as.data.frame(extract(fit_separate, 'ypred_c'))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Original'
temp_df['x'] <- as.numeric(rownames(temp_df))
temp_2 <- rbind(temp_2, temp_df)
temp_2['y_type'] <- 'Prediction'
temp_3 <- rbind(temp_3, temp_2)
temp_3['col_val'] <- 'Combusted'
sep_sens_df <- rbind(sep_sens_df, temp_3)

temp_3 <- as.data.frame(colMeans(as.data.frame(extract(uniform_separate, 'yrep_l'))))
rownames(temp_3) <- NULL
colnames(temp_3) <- c('value')
temp_3['Prior'] <- 'Uniform'
temp_3['x'] <- as.numeric(rownames(temp_3))
temp_df <- as.data.frame(colMeans(as.data.frame(extract(second_separate, 'yrep_l'))))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Second'
temp_df['x'] <- as.numeric(rownames(temp_df))
temp_3 <- rbind(temp_3, temp_df)
temp_df <- as.data.frame(colMeans(as.data.frame(extract(fit_separate, 'yrep_l'))))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Original'
temp_df['x'] <- as.numeric(rownames(temp_df))
temp_3 <- rbind(temp_3, temp_df)
temp_3['y_type'] <- 'Replicate'

temp_2 <- as.data.frame(extract(uniform_separate, 'ypred_l'))
rownames(temp_2) <- NULL
colnames(temp_2) <- c('value')
temp_2['Prior'] <- 'Uniform'
temp_2['x'] <- as.numeric(rownames(temp_2))
temp_df <- as.data.frame(extract(second_separate, 'ypred_l'))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Second'
temp_df['x'] <- as.numeric(rownames(temp_df))
temp_2 <- rbind(temp_2, temp_df)
temp_df <- as.data.frame(extract(fit_separate, 'ypred_l'))
rownames(temp_df) <- NULL
colnames(temp_df) <- c('value')
temp_df['Prior'] <- 'Original'
temp_df['x'] <- as.numeric(rownames(temp_df))
temp_2 <- rbind(temp_2, temp_df)
temp_2['y_type'] <- 'Prediction'
temp_3 <- rbind(temp_3, temp_2)
temp_3['col_val'] <- 'Landfilled'
sep_sens_df <- rbind(sep_sens_df, temp_3)
```

Plot results

```{r hier_sens_plot, fig.height=4, fig.width=5, fig.cap="Sensitivity analysis for hierarchical model."}
ggplot(data=hier_sens_df, aes(x=value, color = Prior)) +
  geom_density() +
  facet_wrap(y_type ~ col_val, scales='free')
```

```{r sep_sens_plot, fig.height=4, fig.width=5, fig.cap="Sensitivity analysis for separate model."}
ggplot(data=sep_sens_df, aes(x=value, color = Prior)) +
  geom_density() +
  facet_wrap(y_type ~ col_val, scales='free')
```

# Model comparison (LOO-CV)

```{r separate_k_plot, fig.height=4, fig.width=5, fig.cap="k-hat values for the separate model."}
#PSIS-LOO for separate model
sep_loo <- rstan::loo(fit_separate)
sep_elpd <- sep_loo$estimates[1,1]
# Effective number of parameters:
sep_p <- sep_loo$estimates[2,1]
plot(sep_loo)
```

```{r hierarchical_k_plot, fig.height=4, fig.width=5, fig.cap="k-hat values for the hierarchical model."}
#PSIS-LOO for hierarchical model
hier_loo <- rstan::loo(fit_hierarchical)
hier_elpd <- hier_loo$estimates[1,1]
# Effective number of parameters:
hier_p <- hier_loo$estimates[2,1]
plot(hier_loo)

```

```{r results='asis'}
loo_table <- rbind(c(hier_elpd, sep_elpd), c(hier_p, sep_p))
colnames(loo_table) <- c('Hierarchical', 'Separate')
rownames(loo_table) <- c('elpd', 'p_eff')
kable(loo_table, caption='LOO-CV results')

```
